---
title: "data exploration case studies"
output: html_notebook
---


# Medicare payment database - subset by state

This notebook can be used to subset the 2014 medicare provider utilization and payment data by state.

The raw data are available [here](http://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/Physician-and-Other-Supplier.html)



This is a rather large file, almost .5Gb in zipped form and over 9 million records.  Using read_tsv (for tab separated values) will produce a roughly 1.8 Gb data frame.

```{r, eval=FALSE}
library(feather)
cms <- readr::read_tsv('data/Medicare_Provider_Util_Payment_PUF_CY2014.txt')
# write_feather(cms, 'data/Medicare_Provider_Util_Payment_PUF_CY2014.feather')
# cms = read_feather('data/Medicare_Provider_Util_Payment_PUF_CY2014.feather')
```

For our purposes we will use the state of Florida (and later Michigan), but first I will note a couple things. Using <span class="func"></span>read_tsv creates an object that is both  <span class="objclass">data.frame</span> and a 'tibble' or  <span class="objclass">tbl</span> object.  Using  <span class="pack">dplyr</span> with such an object will allow for faster operations on larger data.  What you don't want to use are things like  <span class="func">tapply</span> or  <span class="func">aggregate</span>. The former will almost always require additional processing, and the latter is extremely slow.  In addition, if you need additional speed and memory management you can use the  <span class="pack">data.table</span> package. In the following I demonstrate both reading the data and using a group by operation to get the mean charge amount.


```{r dplyrread, message=FALSE, warning=FALSE}
library(dplyr)
system.time({
  cms <- readr::read_tsv('data/Medicare_Provider_Util_Payment_PUF_CY2014.txt', progress=F)
})
```


Now the group operation.

```{r dplyrgroupby, cache=TRUE}
system.time({
cms %>% 
  group_by(nppes_provider_state) %>% 
  summarise(meanCharge = mean(average_submitted_chrg_amt, na.rm=T))
})
```


So around a half second for the grouped summary.  Now we'll do the same with <span class="pack">data.table</span>.

```{r fread, warning=FALSE, message=FALSE}
library(data.table)
system.time({
  cms = fread('data/Medicare_Provider_Util_Payment_PUF_CY2014.txt', verbose=F, showProgress=F)
})
```


```{r datatablegroupby}
system.time({
  cms[,list(meanCharge = mean(average_submitted_chrg_amt, na.rm=T)), by=nppes_provider_state]
})
```

You can ignore the difference in read times. Some times read_tsv was faster, sometimes fread was (the former seemed notably variable). But both were far faster than base R's <span class="func">read.delim</span> (about 4 min) and made better use of memory.  The <span class="objclass">data.table</span> object resulted in a faster grouped operation, and this is to be expected.  The basic trade-off between the two is speed gain for legibility/ease of use.  If you're not using a large data set, <span class="pack">data.table</span> doesn't have much to offer (my opinion!), and if you do, it is probably the clear choice for many operations.  The nice thing is *you don't have to choose*! Both produce objects that are also of class data.frame and so you can use dplyr operations on data.table objects (though the reverse would be more difficult).


Now on to the subset.  One thing you don't want to do in R for this operation is a loop of some kind, e.g. going over each record and if State equals "FL" we keep it. Here we offer three approaches, base R, dplyr and data.table.

```{r subset}
# base R
cms_florida = cms[cms$nppes_provider_state=="FL",]
cms_florida = subset(cms, nppes_provider_state=="FL")

# dplyr
cms_florida = filter(cms, nppes_provider_state=="FL")

# data.table
cms_florida = cms[nppes_provider_state=="FL",]  # assumes conversion to data.table
```


And now we write out the file for future use.  In general you typically will want to only work with data that you have to.  Even in notably smaller data settings, it's just more efficient.

```{r writeout}
readr::write_tsv(cms_florida, 'data/cms_florida.tsv')
```


