---
---

## 


<div style="text-align:center; font-size:200%; font-variant:small-caps;" ><span style='color:#1e90ff'>Data Structures</span></div>

## Data structures

R has several core data structures:

- Vectors
    - Factors
- Lists
- Matrices/arrays
- Data frames


## Vectors

Vectors form the basis of R data structures.

Two main types- atomic and lists, but I will treat lists separately.

Here is an R vector. The *elements* of the vector are numeric values.

```{r vec}
x = c(1, 3, 2, 5, 4)
x
```



## Vectors

All elements of an atomic vector are the same type. Examples include:

- characters
- numeric (double)
- integer
- logical



## Factors

A important type of vector is a <span class="objclass">factor</span>. Factors are used to represent categorical data structures.

```{r factors}
x = factor(1:3, labels=c('q', 'V', 'what the heck?'))
x
```

While the underlying representation is numeric, factors are categorical, and so can't be used as numbers would be.

```{r factors2, error=TRUE}
as.numeric(x)
sum(x)
```



## Matrices

When we move to multiple dimensions, we are dealing with arrays.

Matrices are 2-d arrays, and extremely commonly used.

In R, the vectors making up a matrix *must all be of the same type*.

- e.g. all values in a matrix must be numeric.



## Creating a matrix

Creating a matrix can be done in a variety of ways.

```{r createMatrix, results='hide'}
# create vectors
x = 1:4
y = 5:8
z = 9:12

rbind(x, y, z)   # row bind
cbind(x, y, z)   # column bind
matrix(c(x, y, z), nrow=3, ncol=4, byrow=TRUE)
```


## Lists

Lists in R are highly flexible objects that can contain anything as their elements, even other lists.

- unlike vectors, whose elements must be of the same type. 

Here is an R list. We use the list function to create one. 

```{r}
x = list(1, "apple", list(3, "cat"))
x
```



## Lists

We can use a loop to see the class of each object in the elements. 

```{r list loop}
for(elem in x) class(elem)
```

Lists can be and are often named.

```{r namedlist}
x = list("a" = 25, "b" = -1, "c" = 0)
x["b"]
```

## Data Frames

Data frames are far and away the most commonly used data structure. 

Unlike matrices, they do not have to have the same type of element.  

This is because the `data.frame` class is actually just a list.  

As such, everything about lists applies to data.frames, but they can also be indexed by row or column as well like matrices.

## Creating a data frame

```{r createdf}
mydf = data.frame(a = c(1,5,2),
                  b = c(3,8,1))
```

We can add row names also.

```{r dfrownames}
rownames(mydf) = paste0('row', 1:3)
mydf
```
