---
---
```{r setupggplot, include=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=TRUE, cache=TRUE,
                      R.options=list(width=120), fig.width=4, fig.height=3, fig.align = 'center')
```

## 

<div style="text-align:center;font-variant:small-caps; font-size:200%; color:#1e90ff">ggplot2</div>

## ggplot2

ggplot2 is an extremely popular package for visualization in R.

- and copied in other languages/programs

It entails a <span class="emph">grammar of graphics</span>.

- Every graph is built from the same few parts

Key ideas:

- Aesthetics
- Layers (and geoms)
- Piping
- Facets
- Themes
- Extensions


## ggplot2

Strengths: 

- Ease of getting a good looking plot
- Easy customization
- A lot of data processing is done for you
- Clear syntax
- Easy multidimensional approach
- Equally spaced colors as a default


## Aesthetics

Aesthetics allow one to map data to aesthetic aspects of the plot.

- Size
- Color
- etc.

The function used in ggplot to do this is <span class="func">aes</span>

```{r aes, eval=F}
aes(x=myvar, y=myvar2, color=myvar3, group=g)
```


## Layers

In general, we start with a base layer and add to it.

In most cases you'll start as follows.

```{r layer, eval=FALSE}
ggplot(aes(x=myvar, y=myvar2), data=mydata)
```

This would not produce anything except for a plot background.


## Piping

Layers are added via piping.

The first layers added are typically geoms:

- points
- lines
- density
- text

ggplot2 was using pipes before it was cool, and so it has a different pipe.

Othewise, the concept is the same as before.

```{r pipeplus, eval=FALSE}
ggplot(aes(x=myvar, y=myvar2), data=mydata) +
  geom_point()
```

And now we would have a scatterplot.



## Examples

```{r ggscatter}
library(ggplot2)
data("diamonds"); data('economics')
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point()
```

## Examples

```{r ggline}
ggplot(aes(x=date, y=unemploy), data=economics) +
  geom_line()
```

## Examples

In the following, one setting is not mapped to the data.

```{r ggalpha, fig.width=6, fig.height=4}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(size=carat, color=clarity), alpha=.25) 
```

## Stats

There are many statistical functions built in.

One of the key strengths of ggplot is that you don't have to do much preprocessing.

```{r ggquant}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_quantile()
```


## Stats

```{r ggsmooth}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth()
```


## Stats

```{r ggstatsum}
ggplot(mtcars, aes(cyl, mpg)) + 
  geom_point() +
  stat_summary(fun.data = "mean_cl_boot", colour = "orange", alpha=.75, size = 1)
```


## Facets

Facets allow for panelled display, a very common operation.

In general, we often want comparison plots.

facet_grid will produce a grid.

- Often this is all that's needed

facet_wrap is more flexible.

Both use a formula approach to specify the grouping.

## facet_grid

```{r facetgrid}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
  facet_grid(vs ~ cyl, labeller = label_both)
```


## facet_wrap


```{r facetwrap}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
  facet_wrap(vs ~ cyl, labeller = label_both, ncol=2)
```


## Fine control

<div style='font-size:75%' class='col2'>
ggplot2 makes it easy to get good looking graphs quickly.

However the amount of fine control is extensive.

```{r finecontrolcode, fig.show='hide'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=clarity), alpha=.5) + 
  scale_y_log10(breaks=c(1000,5000,10000)) +
  xlim(0, 10) +
  scale_color_brewer(type='div') +
  facet_wrap(~cut, ncol=3) +
  theme_minimal() +
  theme(axis.ticks.x=element_line(color='darkred'),
        axis.text.x=element_text(angle=-45),
        axis.text.y=element_text(size=20),
        strip.text=element_text(color='forestgreen'),
        strip.background=element_blank(),
        panel.grid.minor=element_line(color='blue'),
        legend.key=element_rect(linetype=4),
        legend.position='bottom')
```
<br><br>
```{r finecontrol, fig.height=6, fig.width=6, echo=FALSE}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=clarity), alpha=.5) + 
  scale_y_log10(breaks=c(1000,5000,10000)) +
  xlim(0, 10) +
  scale_color_brewer(type='div') +
  facet_wrap(~cut, ncol=3) +
  theme_minimal() +
  theme(axis.ticks.x=element_line(color='darkred'),
        axis.text.x=element_text(angle=-45),
        axis.text.y=element_text(size=20),
        strip.text=element_text(color='forestgreen'),
        strip.background=element_blank(),
        panel.grid.minor=element_line(color='blue'),
        legend.key=element_rect(linetype=4),
        legend.position='bottom')
```
</div>


## Themes

In the last example you saw two uses of a theme.

- built-in
- specific customizations

Each argument takes on a specific value or an element function:

- element_rect
- element_line
- element_text
- element_blank


## Themes

The base theme is not too good.

- not for web
- doesn't look good for print either

You will almost invariably need to tweek it.


## Extensions

While many contributed before, ggplot2 now has an extension system.

There is even a website to track the extensions.

- [https://www.ggplot2-exts.org/](- https://www.ggplot2-exts.org/)

Examples include:

- additional themes
- interactivity
- animations
- marginal plots
- network graphs


## Summary ggplot2

ggplot2 is an easy to use, but powerful visualization tool.

Allows one to think in many dimensions for any graph:

- x
- y
- color
- size
- opacity
- facet

2d graphs are not useful for conveying anything but the simplest ideas.

Use ggplot2 to easily go beyond 2d for interesting visualizations.