```{r setupPipes, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=FALSE, cache=FALSE,
                      R.options=list(width=120), fig.width=8, fig.align = 'center')
```


## 

<div style="text-align:center;font-variant:small-caps; font-size:200%; color:#1e90ff">Pipes</div>


##

Note: more detail on much of this part is given in another workshop.





## Pipes

Operators that send what comes before to what comes after.

There are many different pipes.

There are many packages that use their own.

However, the vast majority of packages use the same pipe:



## {.fadeslideslow}

<div style="text-align:center;font-variant:small-caps; font-size:500%; color:#ff5503"> %>% </div>



## Pipes

Here, we'll focus on their use with the dplyr package.

Later, we'll use it for visualizations.

Example.

```{r}
mydf %>% 
  select(var1, var2) %>% 
  filter(var1 == 'Yes') %>% 
  summary
```

Start with a data.frame, select columns from it, filter/subset it, get a summary.

## Pipes for Visualization (more later)

Generic example.

```{r}
basegraph %>% 
  points %>%
  lines %>%
  layout
```



## Pipes

Pipes are best used interactively.

Extremely useful for data exploration.

Common in many visualization packages.

See the <span class="pack">magrittr</span> packge for more pipes.


