---
---

## 

<div style="text-align:center; font-size:200%; font-variant:small-caps;" ><span style='color:#1e90ff'>Vectorization <br>&<br> Apply functions</span></div>


## Base R Indexing Refresher

Slicing vectors

```{r vectorSlice}
letters[4:6]
letters[c(13,10,3)]
```

Slicing matrices/data.frames

```{r matrixSlice, eval=F}
myMatrix[1, 2:3]
```

## Base R Indexing Refresher

Label-based indexing:

```{r dfindexlabel, eval=F}
mydf['row1', 'b']
```

Position-based indexing:

```{r dfindexpos, eval=F}
mydf[1, 2]
```

## Base R Indexing Refresher

Mixed indexing:

```{r dfindexmix, eval=F}
mydf['row1', 2]
```

If the row or column value is empty, all rows or columns are retained.

```{r dfindexslice, eval=F}
mydf['row1',]
mydf[,'b']
```

## Base R Indexing Refresher

Non-contiguous:

```{r dfindexnoncont, eval=F}
mydf[c(1,3),]
```

Boolean:

```{r dfindexbool, eval=FALSE}
mydf[mydf$a >=2,]
```

## Base R Indexing Refresher

List/Data.frame extraction

<span class="func">[</span> : grab a slice of elements/columns

<span class="func">[[</span> : grab specific elements/columns

<span class="func">$</span> : grab specific elements/columns