--- 
title: <span style="font-size:150%; font-variant:small-caps; font-style:italic; color:#1e90ff">webR</span>
author: "Michael Clark"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
    bookdown::tufte_html_book:
      toc: yes
      css: toc.css
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: m-clark/Workshops
description: "An introduction to using R for the web."
---

# Prerequisites
```{r echo=FALSE}
library(tufte)
```

`r newthought('Well known for its statistical capabilities')`, R can also be used for web-scraping, connecting with websites via APIs, html documents and presentations, interactive visualizations, dashboards, and even building entire, possibly interactive, websites.  This talk will provide an overview of web-based use of R. Conceptual introductions, package synopses, and small demonstrations will be presented.


Basic R knowledge is all that is required.  Not even statistical knowledge of any kind is assumerd.


Color coding:


<span class="emph">emphasis</span>

<span class="pack">package</span>

<span class="func">function</span>

<span class="objclass">object/class</span>

<a>link</a>

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

`r margin_note('tufte goes here!')`

